---
title: "Exploratory Data Analysis of MASH Outpatient Procedures"
author: "<a href='https://github.com/ASW-Analyst/'>Andy Wilson</a><br><a href='https://transformationunit.nhs.uk/'>NHS Transformation Unit</a>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
format: 
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    self-contained: true
    css: ../config/nhse_theme.css
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = '100%'
)

library(here)
source(paste0(here(), "/src/requirements/packages.R"))
source(paste0(here(), "/src/requirements/load_packages.R"))
source(paste0(here(), "/src/config/palette.R"))

source(paste0(here(), "/src/scripts/etl/opa_eda_load.R"))
source(paste0(here(), "/src/scripts/etl/eda_dates.R"))
source(paste0(here(), "/src/scripts/processing/eda_op.R"))

source(paste0(here(), "/src/scripts/analysis/eda_op_proc.R"))

source(paste0(here(), "/src/results/tables/eda_op.R"))

```

```{r logo, echo = FALSE}

htmltools::img(src = knitr::image_uri(paste0(here(), "/images/nhse_logo.png")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:5%; padding:10px;',
               width = "180px",
               heigth = "180px")

```

# Introduction
This Exploratory Data Analysis (EDA) focuses on understanding the procedures that have been undertaken for patients with a MASH diagnosis. Similarly to the inpatients analysis this is classified as a primary diagnosis recorded at outpatient attendance of:

- **K760**: _Fatty (change of) liver, not elsewhere classified_
- **K740**: _Hepatic Fibrosis_

This analysis has explored appointments at any provider across England between **`r format(as.Date(dates_in_where_clause_op[1]), "%d %B %Y")`** and **`r format(as.Date(dates_in_where_clause_op[2]), "%d %B %Y")`**.

# Procedures

::: {.callout-tip}
## Key findings

- `r round(opa_types[[3,3]] *100, 2)`% of appointments recorded a procedure.
- Ultrasound Elastography was the most common procedure undertaken at outpatient appointment.
:::

There was a total of `r comma(sum(opa_types$Total))` appointments that were attended by a patient diagnoses with MASH. Of these `r round(opa_types[[2,3]] *100, 2)`% had no procedure codes recorded. Therefore, it is difficult to draw robust conclusions as to whether any of these appointments had a procedure. `r round(opa_types[[3,3]] *100, 2)`% of appointments had an OPCS code recorded that confirmed some form of procedure undertaken.

The table below shows a the volume of each procedure undertaken in an outpatient setting for those with more than 10 procedures recorded in total:

```{r procs table}

table_eda_op_procs

```